---
title: "Game of thrones"
author: "Upaluk"
date: "updated `r format(Sys.time(), '%d %B, %Y')` "
output: 
  prettydoc::html_pretty:
  theme: cayman
  toc: yes
toc_float: true
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)

library(tidyverse)
library(here)

# For text mining:

library(pdftools)
library(tidytext)
library(textdata) 
library(ggwordcloud)
```

```{r get-document}
got_path <- here("data","got.pdf")
got_text <- pdf_text(got_path)
```

Example: This is page 9 (e.g. Page 9)? 
```{r single-page}
got_p9 <- got_text[9]
got_p9
```
### Some wrangling:

- Split up pages into separate lines (separated by `\n`) using `stringr::str_split()`
- Unnest into regular columns using `tidyr::unnest()`
- Remove leading/trailing white space with `stringr::str_trim()`

```{r split-lines}
got_df <- data.frame(got_text) %>% 
  mutate(text_full = str_split(got_text, pattern = '\n')) %>% 
  unnest(text_full) %>% 
  mutate(text_full = str_trim(text_full)) 

# More information: https://cran.r-project.org/web/packages/stringr/vignettes/regular-expressions.html
got_df
```

Now each line, on each page, is its own row, with extra starting & trailing spaces removed. 

### Get the tokens (individual words) in tidy format

Use `tidytext::unnest_tokens()` (which pulls from the `tokenizer`) package, to split columns into tokens. We are interested in *words*, so that's the token we'll use:

```{r tokenize}
got_tokens <- got_df %>% 
  unnest_tokens(word, text_full)

# See how this differs from `ipcc_df`
# Each word has its own row!
got_tokens
```

Let's count the words!

```{r count-words}
got_wc <- got_tokens %>% 
  count(word) %>% 
  arrange(-n)

got_wc
```

### Remove stop words:

See `?stop_words` and `View(stop_words)`to look at documentation for stop words lexicons.

We will *remove* stop words using `tidyr::anti_join()`:

```{r stopwords}
got_stop <- got_tokens %>% 
  anti_join(stop_words) %>% 
  select(-got_text)

got_stop
```


Now check the counts again: 

```{r count-words2}
got_swc <- got_stop %>% 
  count(word) %>% 
  arrange(-n)

got_swc
```
What if we want to get rid of all the numbers (non-text) in `got_stop`?

```{r skip-numbers}
# This code will filter out numbers by asking:
# If you convert to as.numeric, is it NA (meaning those words)?
# If it IS NA (is.na), then keep it (so all words are kept)
# Anything that is converted to a number is removed

got_no_numeric <- got_stop %>% 
  filter(is.na(as.numeric(word)))
```

### A word cloud of IPCC report words (non-numeric)

See more: https://cran.r-project.org/web/packages/ggwordcloud/vignettes/ggwordcloud.html

```{r wordcloud-prep}
# There are almost 2000 unique words 
length(unique(got_no_numeric$word))

# We probably don't want to include them all in a word cloud. Let's filter to only include the top 100 most frequent?
got_top100 <- got_no_numeric %>% 
  count(word) %>% 
  arrange(-n) %>% 
  head(100)
got_top100
```


```{r wordcloud}
got_cloud <- ggplot(data = got_top100, aes(label = word)) +
  geom_text_wordcloud() +
  theme_minimal()

got_cloud
```

That's underwhelming. Let's customize it a bit:

```{r wordcloud-pro}
ggplot(data = got_top100, aes(label = word, size = n)) +
  geom_text_wordcloud_area(aes(color = n), shape = "star") +
  scale_size_area(max_size = 20) +
  scale_color_gradientn(colors = c("purple","red","pink")) +
  theme_minimal()
```

"afinn": Words ranked from -5 (very negative) to +5 (very positive)

```{r afinn}
get_sentiments(lexicon = "afinn")
# You should be prompted to install lexicon - choose yes!


# Let's look at the pretty positive words:
afinn_pos <- get_sentiments("afinn") %>% 
  filter(value %in% c(3,4,5))

# Do not look at negative words in class. 
afinn_pos
```

bing: binary, "positive" or "negative"

```{r bing}
get_sentiments(lexicon = "bing")
# You should be prompted to install lexicon - choose yes!

```

nrc:https://saifmohammad.com/WebPages/NRC-Emotion-Lexicon.htm
Includes bins for 8 emotions (anger, anticipation, disgust, fear, joy, sadness, surprise, trust) and positive / negative. 

**Citation for NRC lexicon**: Crowdsourcing a Word-Emotion Association Lexicon, Saif Mohammad and Peter Turney, Computational Intelligence, 29 (3), 436-465, 2013.

Now nrc:

```{r nrc}
get_sentiments(lexicon = "nrc")
# You should be prompted to install lexicon - choose yes!

```



Let's do sentiment analysis on the got text data using afinn, and nrc. 


### Sentiment analysis with afinn: 

First, bind words in `got_stop` to `afinn` lexicon:

```{r bind-afinn}
got_afinn <- got_stop %>% 
  inner_join(get_sentiments("afinn"))
got_afinn
```
Figure 1
Let's find some counts (by sentiment ranking):
```{r count-afinn}
got_afinn_hist <- got_afinn %>% 
  count(value)

# Plot them: 
ggplot(data = got_afinn_hist, aes(x = value, y = n)) +
  geom_col(aes(fill = value)) +
  theme_bw()
```

Investigate some of the words in a bit more depth:

```{r afinn-2}
# What are these '2' words?
got_afinn2 <- got_afinn %>% 
  filter(value == 2)


```

Figure 2
```{r afinn-2-more}
# Check the unique 2-score words:
unique(got_afinn2$word)

# Count & plot them
got_afinn2_n <- got_afinn2 %>% 
  count(word, sort = TRUE) %>% 
  mutate(word = fct_reorder(factor(word), n))

# comment I have chosen only to show the top 25 words, to make it easier to read. 

got_afinn2_n %>%
  slice_max(n, n = 25) %>%
  ggplot(aes(x = reorder(word, n), y = n)) +
  geom_col() +
  coord_flip() +
  theme_bw()

 
```


Or we can summarize sentiment for the book:

```{r summarize-afinn}
got_summary <- got_afinn %>% 
  summarize(
    mean_score = mean(value),
    median_score = median(value)
  )
got_summary 
```

We can use the NRC lexicon to start "binning" text by the feelings they're typically associated with. As above, we'll use inner_join() to combine the IPCC non-stopword text with the nrc lexicon:

```{r bind-bing}
got_nrc <- got_stop %>% 
  inner_join(get_sentiments("nrc"))
```

Wait, won't that exclude some of the words in our text? YES! We should check which are excluded using `anti_join()`:

```{r check-exclusions}
got_exclude <- got_stop %>% 
  anti_join(get_sentiments("nrc"))

# View(got_exclude)

# Count to find the most excluded:
got_exclude_n <- got_exclude %>% 
  count(word, sort = TRUE)

head(got_exclude_n)
```

Now find some counts:
Figure 3
```{r count-bing}
got_nrc_n <- got_nrc %>% 
  count(sentiment, sort = TRUE)

# And plot them:

ggplot(data = got_nrc_n, aes(x = sentiment, y = n)) +
  geom_col(aes(fill = sentiment))+
  theme_bw()
```


Or count by sentiment *and* word, then facet:
Figure 4 
```{r count-nrc}
got_nrc_n5 <- got_nrc %>% 
  count(word,sentiment, sort = TRUE) %>% 
  group_by(sentiment) %>% 
  top_n(5) %>% 
  ungroup()

got_nrc_gg <- ggplot(data = got_nrc_n5, aes(x = reorder(word,n), y = n, fill = sentiment)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~sentiment, ncol = 2, scales = "free") +
  coord_flip() +
  theme_minimal() +
  labs(x = "Word", y = "count")

# Show it
got_nrc_gg

# Save it
ggsave(plot = got_nrc_gg, 
       here("figures","got_nrc_sentiment.png"), 
       height = 8, 
       width = 5)

```

Wait, so "stone" is showing up in NRC lexicon as "anger"? Let's check:
```{r nrc-stone}
ston <- get_sentiments(lexicon = "nrc") %>% 
  filter(word == "stone")

# Yep, check it out:
ston
```

# What are the most common meaningful words and what emotions do you expect will dominate this volume? Are there any terms that are similarly ambiguous to the 'confidence' above? 

The most common meaningful words in the chart highlight a world driven by power, family, and mortality. Terms like lord, king, brother, mother, and child emphasize the importance of lineage and relationships, while death, die, dark, and fire reinforce the constant presence of danger. Words such as time and watch suggest inevitability and tension building toward future conflict.

From this, the dominant emotions are clearly fear, anticipation, sadness, and anger. Fear is reflected in repeated mentions of death, fire, and watch, suggesting vigilance and threat. Anticipation, led by time, conveys a sense that major events are always approaching. Sadness appears through leave, fell, and dark, indicating loss and emotional weight. Anger emerges in harsher language like bastard and conflict-related terms. However, emotions like trust and joy persist through words such as mother, child, and smile, hinting at human connection amid chaos.

Several words are emotionally ambiguous, similar to “confidence.” Watch can imply protection or danger, fire can symbolize destruction or strength, and sweet may be sincere or ironic. Even lord spans multiple emotional categories, making it context-dependent.

Overall, the sentiment suggests a harsh, unstable world balanced by moments of loyalty and connection.


### Credits: 
This tutorial is inspired by Allison Horst's Advanced Statistics and Data Analysis.



