---
title: "Are we more civilized today?"
author: "Upaluk"
date: " Created in 2021, updated `r format(Sys.time(), '%d %B, %Y')` "
output: 
  prettydoc::html_pretty:
  theme: cayman
  toc: yes
toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message=FALSE)
```
#  Explore Homicide rates in NW Europe
In this exercise, you will load a Manuel Eisner's dataset that captures the developments in homicides from Middle Ages to 20th century. 

**Your task: Explore the data and create visualisations as instructed in the script.** 

- **Upload to Github this .Rmd file with solutions and its .html knitted version.**  
- **Submit to Brightspace a document that follows the template and contains (1) URL of this script and its knitted version in Github, (2) one plot of your choice,  and (3) answer of 250 words to the Final Question.**



This exercise is based on the dataset provided by [OurWorldInData project](https://ourworldindata.org/homicides) based at the Oxford University.

## The long-term trend in Homicides in Western Europe

Understanding how homicide rates have changed prior to the modern era requires the help of historians and archivists. Manuel Eisner, a criminology professor at the University of Cambridge, and his colleagues published the Historical Violence Database \href{https://cjrc.osu.edu/research/interdisciplinary/hvd}: a compilation of data on long-term trends in homicide rates, in addition to qualitative information such as the cause of death, perpetrator and victim. This database is limited to countries with relatively complete historical records on violence and crime – mainly Western Europe and the US.

Starting in the second half of the nineteenth century, these European regions have consistent police records of those accused of murder or manslaughter and annual counts of homicide victims. To go back further in time, reaching as far back as the thirteenth century, Eisner collected estimates (from historical records of coroner reports, court trials, and the police) of homicide rates made in over ninety publications by scholars.

Homicide rates – measured as the number of homicides per 100,000 individuals – up to 1990 are sourced from Eisner’s (2003) publication and the Historical Violence Database. 

Are homicide rates in Europe today lower or higher than in the past? Using the provided dataset, display and describe the long-run homicide rates for the five European regions: Italy, England, Germany, Netherlands and Scandinavia.

```{r library}
library(tidyverse)
```

## Load the available data from ourworldindata.org

You should always interrogate the source of your data. Who compiled it, from where, what is missing, how representative the data are? Check the data/Metadata.txt to learn about the data provenance.

```{r load-data}
Western_Europe <- read_csv("data/homicide-rates-across-western-europe.csv")
```

## Inspect the data

How clean and analysis-ready is the dataset? Do you understand what the column names represent? What is the difference between rate and homicide number?

```{r inspect}
head(Western_Europe)
```

Ok, the data look good except for the column `Homicide rate in Europe over long-term (per 100,000)` which is not very easy to work with.

- Use the `names()` function and assignment key to relabel this column to `homicides_per_100k` 


```{r relabel column}
# names() function produces the names of columns as a character vector with 4 elements
names(Western_Europe)
names(Western_Europe)[4]

# If we can select the fourth element, we can also rename it / reassign a name to it
names(Western_Europe)[4] <- "Homicides_per_100k"
names(Western_Europe)[4]
# check the new name!
names(Western_Europe)
```

Now, that you have looked at what the data looks like and what it represents, and streamlined it, let's see what big picture it contains.

## Let's see what the long-term trend is in homicides

- use `ggplot()` function and remember the`+` at the end of the line
- chose a meaningful `geom_......()`  for geometry (hint: points are not great)
- load `Year` on the `x` axis and `homicides_per_100k` column in y axis
- to color individual country entries consistently, assign the country column to the argument `color`.
- provide meaningful title and axis labels
- remember to change the `eval` flag so that the code chunk renders when knitted 

```{r plot-homicides, eval = TRUE}

# option 2 - this one starts with the dataset and continues with ggplot()

Western_Europe %>% 
  ggplot(aes(x = Year, y = Homicides_per_100k, colour = Entity))+
  geom_line(lwd = 2) +
  theme_classic()+
  labs(y= "Homicide per 100,000 people",
       title = "Historical homicides")+
  theme(text = element_text(size = 18))
  
ggsave("Homicides.png", width = 16)
```
Alright, the homicide rates should all be descending over time. What a comfort. But the viz is not super clear. Let's check the rates for individual countries.

## Uncouple the homicides of individual countries for easier view

You can visualize each country's trend separately by adding an extra argument to the ggplot, the `facet_wrap()` and feeding it the country column. If in doubt, check your ggplot tutorial and your country column name for exact usage.

- reuse the ggplot from the chunk above
- insert `facet_wrap()` after the specification of geometry to split countries in separate charts
- change the facet "layout" to two columns and three rows so that the trends are easier to see in horizontal layout. 


```{r Historical homicides, eval = TRUE}
Western_Europe %>% 
  ggplot(aes(x = Year, y = Homicides_per_100k, colour = Entity))+
  geom_line(lwd = 2) +
  theme_classic()+
  labs(y= "Homicide per 100,000 people",
       title = "Historical homicides",
       colour = "Country")+
  theme(text = element_text(size = 18),
        legend.position = "bottom")+
  facet_wrap(~ Entity)

ggsave("HomicidesFacet.png", width = 16)
  
```

## Compare the trends in homicide with the pattern of reign duration among Danish rulers through time.

- Load your Danish king dataset. Hopefully it is tidy and your years and duration of reign are all numeric.
- You need to have a consistent way of plotting the rulers' reign on the x axis, so I recommend you create a midyear column by calculating the middle of each monarch's rule (Hint: `midyear` = `endyear` - (`endyear`-`startyear`)/2)
- Start a ggplot plotting midyear on x axis and duration on y axis
- Try `geom_smooth()` for geometry 
- Provide meaningful labels and a title
- How would you characterize the trend compared to the homicides above?
The less homicides there were, the longer danish monarchs ruled. 

```{r kings,eval = TRUE}
# YOUR CODE HERE:
kings <- read_csv2("data/Monarker.csv")
glimpse(kings)

kings <- kings %>% 
  mutate(duration = Reign_End - Reign_start)

kings <- kings %>%
  mutate(midyear = Reign_End - (Reign_End - Reign_start)/2)
```

Here we start with ggplot
```{r midyear, eval = TRUE}

kings %>% 
  filter(
    !is.na(midyear),
    !is.na(duration),
    !is.na(`King/Queen`)) %>%
  ggplot(aes(x = midyear, y = duration))+
  geom_point() +
  geom_smooth() +
  theme_minimal()+
  labs(y= "duration",
       title = "danish monarchs reign duration",
       x = "Year")+
  theme(text = element_text(size = 18))

ggsave("KingAndQueens.png")

```


##Final tasks:

1) Plot: In the facetted plot above, move the legend from the current position on the side to below the facets, and label it "Country" instead of "Entity".

2) Rmarkdown: 

  - edit the author of the document, and convert 'Final Tasks' into heading #2 (like the other headings) 
  - add a `floating table of contents` to your Rmarkdown document, 
  - provide informative `chunk-names` and edit flags in your R chunks to `TRUE`, and
  - automatically generate a `timestamp` to show when the document was last updated. (Hint: check the Rmarkdown episode in our Data Carpentry tutorial)

###  Final question: 

*In 250 words articulate your answer on the basis of the data visualisations to the following question*: **are we more civilized today?** 

The visualisations show a clear, long-term decline in homicide rates across all regions from the medieval period to the present. In places like England, Italy, and the Netherlands/Belgium, rates fall dramatically from very high levels (often above 40–60 per 100,000) to near zero today. This suggests a substantial reduction in interpersonal violence over time.

On this basis, one could argue that societies have become more “civilized,” if civilization is associated with greater self-control, stronger legal institutions, and reduced tolerance for violence. The consistency of the downward trend across multiple regions strengthens this interpretation, pointing to broad structural changes such as the development of states, policing, and cultural norms discouraging violence.

However, the data only capture one dimension of “civilization”—homicide. They do not account for other forms of harm, such as war, structural inequality, or non-lethal violence. It is also possible that improvements in record-keeping affect earlier estimates.

Overall, the charts provide strong evidence that deadly interpersonal violence has decreased significantly over time, supporting the argument that, in this specific sense, we are more civilized today.
